---
title: "Functional Programming"
author: "Jan Vitek"
date: "Wednesday May 3, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```
- Functionals: `lapply` and family
- Example: better solution to problem using `lapply`
- More examples:
    + `sapply`
    + `tapply`
    + `vapply`
    + `apply`
    + etc.
- Anonymous functions
- Closures
- When to use a loop vs. functional
````

*The material in this lecture is based on Hadley Wickham's Advanced R book.*


Looking back to a previous example.

```{r}
set.seed(1014)
df <- data.frame(replicate(6, sample(c(1:10, -99), 6, rep = TRUE)))
names(df) <- letters[1:6]
fix_missing <- function(x) {
  x[x == -99] <- NA
  x
}
df$a <- fix_missing(df$a)
df$b <- fix_missing(df$b)
df$c <- fix_missing(df$c)
df$d <- fix_missing(df$d)
df$e <- fix_missing(df$e)
df$f <- fix_missing(df$e)
```

`lapply()` takes three inputs: x, a list; f, a function; and ..., other arguments to pass to `f()`. It applies the function to each element of the list and returns a new list. `lapply(x, f, ...)` is equivalent to the following for loop:

```
out <- vector("list", length(x))
for (i in seq_along(x)) 
  out[[i]] <- f(x[[i]], ...)
```

`lapply()` is called a functional, because it takes a function as an argument. 

Data frames are lists. We just need to make sure we get back a data frame, not a list. Instead of assigning the results of lapply() to df, we’ll assign them to df[]. R’s usual rules ensure that we get a data frame, not a list. 

```{r}
fix_missing <- function(x) {
  x[x == -99] <- NA
  x
}
df[] <- lapply(df, fix_missing)
```

And we can work on portions of the input

```{r}
df[1:5] <- lapply(df[1:5], fix_missing)
```

What if different columns used different codes for missing values? You might be tempted to copy-and-paste:

```{r}
fix_missing_99 <- function(x) {
  x[x == -99] <- NA
  x
}
fix_missing_999 <- function(x) {
  x[x == -999] <- NA
  x
}
fix_missing_9999 <- function(x) {
  x[x == -999] <- NA
  x
}
```

As before, it’s easy to create bugs. Instead we could use closures, functions that make and return functions. 

```{r}
missing_fixer <- function(na_value) 
  function(x) { x[x==na_value]<-NA; x }

fix_missing_99 <- missing_fixer(-99)
fix_missing_999 <- missing_fixer(-999)
fix_missing_99(c(-99, -999))
```

# Functionals

A higher-order function is a function that takes a function as an input or returns a function as output. 

# My first functional: lapply()

lapply rougly works like this:

```{r}
lapply2 <- function(x, f, ...) {
  out <- vector("list", length(x))
  for (i in seq_along(x)) 
    out[[i]] <- f(x[[i]], ...)
  out
}
```

`lapply()` is a wrapper for a common for loop pattern: create a container for output, apply `f()` to each component of a list, and fill the container with the results.

```{r,eval=F}
# Create some random data
l <- replicate(20, runif(sample(1:10, 1)), simplify = FALSE)

# With a for loop
out <- vector("list", length(l))
for (i in seq_along(l)) out[[i]] <- length(l[[i]])
unlist(out)
##  [1]  3  9  1  5  6  9 10  3  3  1  8  3  3  4 10  8  6  8  7  3

# With lapply
unlist(lapply(l, length))
##  [1]  3  9  1  5  6  9 10  3  3  1  8  3  3  4 10  8  6  8  7  3
```

`lapply()` is also useful when you want to do something to each column of a data frame:

```{r}
# What class is each column?
unlist(lapply(mtcars, class))
# Divide each column by the mean
mtcars[] <- lapply(mtcars, function(x) x / mean(x))
```

There are three basic ways to use lapply():

 * `lapply(xs, function(x) {})`
 * `lapply(seq_along(xs), function(i) {})` 
 * `lapply(names(xs), function(nm) {})`
 
Typically you’d use the first form because `lapply()` takes care of saving the output for you.

# Vector output: sapply and vapply

`sapply()` and `vapply()` are similar to `lapply()` but produce an atomic vector. While sapply() guesses, vapply() takes an argument specifying the output type.

```{r}
sapply(mtcars, is.numeric)
##  mpg  cyl disp   hp drat   wt qsec   vs   am gear carb 
## TRUE TRUE TRUE TRUE TRUE TRUE TRUE TRUE TRUE TRUE TRUE
vapply(mtcars, is.numeric, logical(1))
##  mpg  cyl disp   hp drat   wt qsec   vs   am gear carb 
## TRUE TRUE TRUE TRUE TRUE TRUE TRUE TRUE TRUE TRUE TRUE
```

If the function returns results of different types or lengths, `sapply()` will silently return a list, while `vapply()` will throw an error. 

# Multiple inputs: Map (and mapply)

With `lapply()`, only one argument to the function varies. This makes it poorly suited for some problems. For example, how would you find a weighted mean when you have two lists, one of observations and the other of weights?

```{r}
# Generate some sample data
xs <- replicate(5, runif(10), simplify = FALSE)
ws <- replicate(5, rpois(10, 5) + 1, simplify = FALSE)
```

It’s easy to use `lapply()` to compute the unweighted means:

```{r}
unlist(lapply(xs, mean))
## [1] 0.5001362 0.5957105 0.5971894 0.5381643 0.5704373
```

But how could we supply the weights to `weighted.mean()`? `lapply(x, means, w)` won’t work because the additional arguments to `lapply()` are passed to every call. We could change looping forms:

```{r}
unlist(lapply(seq_along(xs), function(i) {
  weighted.mean(xs[[i]], ws[[i]])
}))
## [1] 0.5096482 0.5621915 0.5599182 0.4842896 0.5773232
```

A cleaner alternative is to use Map:

```{r}
unlist(Map(weighted.mean, xs, ws))
## [1] 0.5096482 0.5621915 0.5599182 0.4842896 0.5773232
```

This is equivalent to:

```{r}
stopifnot(length(xs) == length(ws))
out <- vector("list", length(xs))
for (i in seq_along(xs)) {
  out[[i]] <- weighted.mean(xs[[i]], ws[[i]])
}
```


Map is useful whenever you have multiple lists/df to process in parallel. E.g. to standardise columns one can compute the means and then divide by them. 

```{r}
mtmeans <- lapply(mtcars, mean)
mtmeans[] <- Map(`/`, mtcars, mtmeans)
```

If some of the arguments should be fixed (e.g. na.rm), use an anonymous function:

```
Map(function(x, w) weighted.mean(x, w, na.rm = TRUE), xs, ws)
```
 
# Manipulating lists

Another way of thinking about functionals is as a set of general tools for altering, subsetting, and collapsing lists. Every functional programming language has three tools for this: `Map()`, `Reduce()`, and `Filter()`. 

## Reduce()

`Reduce()` reduces a vector to a scalar by recursively calling a function, f, two arguments at a time. It combines the first two elements with f, then combines the result of that call with the third element, and so on. `Reduce(f, 1:3) == f(f(1, 2), 3)`. 

```{r}
Reduce(`+`, 1:3) # -> ((1 + 2) + 3)
Reduce(sum, 1:3) # -> sum(sum(1, 2), 3)
```

The essence of `Reduce()` is:

```
Reduce2 <- function(f, x) {
  out <- x[[1]]
  for(i in seq(2, length(x)))
    out <- f(out, x[[i]])
  out
}
```

`Reduce()` is useful for implementing many types of recursive operations, like merges and intersections. Imagine you have a list of numeric vectors, and you want to find the values that occur in every element:

```{r}
l <- replicate(5, sample(1:10, 15, replace = T), simplify = FALSE)
str(l)
## List of 5
##  $ : int [1:15] 3 5 4 7 8 1 10 2 10 3 ...
##  $ : int [1:15] 9 4 9 6 4 9 4 3 3 8 ...
##  $ : int [1:15] 6 4 4 9 7 6 9 4 9 7 ...
##  $ : int [1:15] 5 5 8 5 7 4 7 7 1 10 ...
##  $ : int [1:15] 2 10 9 7 5 1 4 6 6 3 ...
```

You could do that by intersecting each element in turn:

```{r}
intersect(intersect(intersect(intersect(l[[1]], l[[2]]),
  l[[3]]), l[[4]]), l[[5]])
## [1] 3 4 7 8
```
That’s hard to read. With `Reduce()`, the equivalent is:

```{r}
Reduce(intersect, l)
## [1] 3 4 7 8
```

## Predicate functionals

A predicate is a function that returns a single TRUE or FALSE, like `is.character`, `all`, or `is.NULL`. A predicate functional applies a predicate to each element of a list or data frame. There are three useful predicate functionals in base R: `Filter()`, `Find()`, and `Position()`.

`Filter()` selects only those elements which match the predicate.

`Find()` returns the first element which matches the predicate.

`Position()` returns the position of the first element that matches the predicate.

Another useful predicate functional is `where()`, a custom functional that generates a logical vector from a list and a predicate:

```{r}
where <- function(f, x) 
  vapply(x, f, logical(1))
```



```{r}
df <- data.frame(x = 1:3, y = c("a", "b", "c"))
where(is.factor, df)
##     x     y 
## FALSE  TRUE
str(Filter(is.factor, df))
## 'data.frame':    3 obs. of  1 variable:
##  $ y: Factor w/ 3 levels "a","b","c": 1 2 3
str(Find(is.factor, df))
##  Factor w/ 3 levels "a","b","c": 1 2 3
Position(is.factor, df)
## [1] 2
```


