---
title: "May institute Data visualization module: Hands-on 2"
author: "Ting Huang and Dan Guo"
date: "May 8, 2017"
output: 
  html_document:
    self_contained: true
    toc: true
    toc_float: true
    fig_caption: no	
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Objectives
* Data Exploration
* Plot types
    + Bar plot and pie chart
    + Histogram
    + Scatter plot
    + Text plot
    + Boxplot and violin plot
    + Line chart

---    

# 1. Data Exploration
In this section we’ll continue using the colonrectal cancer dataset. But besides sub group information, we'll explore other sample information.

```{r, echo=T, warning=FALSE}
CRC <- read.csv("./data/CRC_train.csv")
str(CRC[,73:79])

# Inspect the possible values for the annotation columns
unique(CRC[, 'Cancer_stage'])
unique(CRC[, 'Tumour_location'])
unique(CRC[, 'Sub_group'])
```

---

# 2. Plot types
## 2.1 Bar plot and pie chart
`geom_bar()` makes the height of the bar proportional to the number of cases in each group.
```{r,warning=FALSE, message=F}
library(ggplot2)
# Investigate the number of each sub_group
g1 <- ggplot(CRC, aes(Sub_group)) 
g1 + geom_bar()
# horizontal barplot
g1 + geom_bar() + 
  coord_flip()
```

We can also calculate the number of patients in each group manually and then make bar plot with `stat="identity"`
```{r, warning=FALSE, message=F}
library(dplyr)
# count the number of samples in each sub group
n_subgroup <- CRC %>% group_by(Sub_group) %>% summarise(N = n())
n_subgroup
ggplot(n_subgroup, aes(x=Sub_group, y=N)) + 
  geom_bar(stat="identity")
```


We want to know the count of different gender patients in each sub group
```{r}
g2 <- ggplot(CRC, aes(Sub_group, fill=factor(Gender)))
# Stacked barplot puts overlapping bars on top of each other
g2 + geom_bar(position = "stack")
# place overlapping bars side-by-side
g2 + geom_bar(position = "dodge")
#displays proportions
g2 + geom_bar(position = "fill") 
```

Pie chart is changed from bar chart through `coord_polar()`
```{r}
ggplot(data = CRC) + 
  geom_bar(mapping = aes(x = factor(1), fill = factor(Sub_group))) + 
  coord_polar(theta = "y")
```


## 2.2 Histogram

We want to Investigate the abundance distribution of one protein. `geom_histogram()` shows the distribution of a single variable
```{r}
g3 <- ggplot(CRC, aes(SERPINA3))
g3 + geom_histogram()
```

`freqency polygons` show the distribution of a single variable with lines
```{r}
g3 + geom_freqpoly()
```


> **Challenge**
> If you want to know the density distribution of protein AFM among all the observations, what plot will you make? 

$~$

$~$

$~$

$~$

$~$

$~$

$~$
$~$

```{r}
g <- ggplot(CRC, aes(AFM))
g + geom_histogram()
g + geom_freqpoly()
```

## 2.3 Scatter plot
We want to investigate the relation between two proteins. `geom_point()` is used to create scatterplots. The scatterplot is most useful for displaying the relationship between two continuous variables.
```{r}
g <- ggplot(CRC, aes(x = SERPINA3, y = TIMP1))
g + geom_point()
```

If you have a scatterplot with a lot of noise, it can be hard to see the dominant pattern. In this case it’s useful to add a smoothed line to the plot with `geom_smooth()`
```{r}
g + geom_point() + 
  geom_smooth()
```

`smoothScatter()` produces a smoothed color density representation of a scatterplot
```{r}
smoothScatter(CRC$SERPINA3, CRC$TIMP1)
```


> **Challenge**
> If you want to the relationship of abundance between protein AFM and AHSG, what plot will you make?

$~$

$~$

$~$

$~$

$~$

$~$

$~$
$~$

```{r}
ggplot(CRC, aes(x = AFM, y = AHSG))+
  geom_point()+
  geom_smooth()
```

## 2.4 Text plot
`geom_text` can add text to the plot
```{r}
g1 <- ggplot(CRC, aes(x = Sample, y = TIMP1))
g1 + geom_text(aes(label = Age))
```

`check_overlap` can remove the overlapped texts
```{r}
g1 + geom_text(aes(label = Age), check_overlap = TRUE)
```

We can aslo add points to the plot.
```{r}
g1 + geom_text(aes(label = Age), check_overlap = TRUE) +
  geom_point()
```

`nudge_x` and `nudge_y` can offset the text
```{r}
g1 + geom_text(aes(label = Age), nudge_y = 0.05, check_overlap = TRUE) +
  geom_point(shape = 1, aes(color = Sub_group))
```

ggrepel implements functions to repel overlapping text labels away from each other and away from the data points that they label. 
```{r}
library(ggrepel)
g1 + geom_point(shape = 1, aes(color = Sub_group)) +
  geom_text_repel( aes( label = Age),  check_overlap = F, 
             position=position_jitter(width=.1, height=.2)  )
```


Adding labels to bar plot
```{r}
# create a ggplot data
ggplot(n_subgroup, aes(x=Sub_group, y=N)) +
  # draw the bar plot
  geom_bar(stat="identity") +
  # create the number text above the bar in white
  geom_text(aes(label=N), vjust=1.5, colour="white")
```
 
## 2.5 Boxplot and violin plot
We want to investigate how protein abundance varies across different groups. `geom_boxplot()` summarise the shape of the distribution with a handful of summary statistics.
```{r}
b <- ggplot(CRC, aes(Sub_group, SERPINA3)) 
b + geom_boxplot()
```

We can also overlay multiple geoms by simply adding them one after the other.
```{r}
b + geom_boxplot() + geom_jitter() 
b + geom_jitter() + geom_boxplot()
b + geom_jitter(alpha = 0.5) + geom_boxplot()
```

We want to consider another factor *gender*
```{r}
b + geom_boxplot(aes(fill= factor(Gender)), position = "dodge")
```

A violin plot is a mirrored density plot displayed in the same way as a boxplot.
```{r}
b + geom_violin()
```

## 2.6 Line chart
We want to know how the protein abundance changes across different samples.
```{r}
ggplot(CRC[1:100,],aes(Sample, SERPINA3, group = 1)) + 
  geom_point(size = 0.5) +
  geom_line() 
```


> **Challenge**
> If you want to compare the abundance distribution of protein AFM among different groups, what plot will make?
> And you also want to see the difference between  male and female in each group, what change will make?
> How to make the boxplot horizontal direction?
> Add data points to the boxplot.
$~$

$~$

$~$

$~$

$~$

$~$

$~$
$~$


```{r}
ggplot(CRC,aes(Sub_group, AFM)) + 
  geom_boxplot()

ggplot(CRC,aes(Sub_group, AFM, fill= factor(Gender))) + 
  geom_boxplot(position = "dodge")

ggplot(CRC,aes(Sub_group, AFM, fill= factor(Gender))) + 
geom_boxplot(position = "dodge") +
  coord_flip()

ggplot(CRC,aes(Sub_group, AFM, fill= factor(Gender))) + 
  geom_boxplot(position = "dodge")+
  geom_jitter()
```