---
title: "May institute Data visualization module: Hands-on 2"
author: "Ting Huang"
date: "May 9, 2017"
output: 
  html_document:
    self_contained: true
    toc: true
    toc_float: true
    fig_caption: no	
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Summary
* Creating a new RStudio project
* Reading in data in R
* Improve figure clarity
    + Deal with overplotting
    + Order bar chart
    + Zooming
    + Change labels, themes, and scales
    
http://sebastianraschka.com/Articles/heatmaps_in_r.html

---


# 1. Reading in data
In this section we’ll continue using mpg and CRC dataset.

```{r, echo=T, warning=FALSE, message=FALSE}
library(Biobase)

################################################################
# Read data
################################################################
CRC <- read.csv("./data/CRC_train.csv")
CRC.prot <- CRC[,1:72]
CRC.anno <- CRC[,73:79]

# Deal with missing value
# First option: remove the samples with missing values
dim(na.omit(CRC.prot))

# Second option: impute the missing values
random.imp <- function (a){
  missing <- is.na(a)
  n.missing <- sum(missing)
  a.obs <- a[!missing]
  imputed <- a
  # imputed[missing] <- sample(a.obs, n.missing, replace=TRUE)
  imputed[missing] <- median(a.obs)
  # imputed[missing] <- 0
  return (imputed)
}
pMiss <- function(x){sum(is.na(x))/length(x)*100}
samplemissing <- apply(CRC.prot,1,pMiss)
# Only keep the samples with less than 5% missing values
selectedSamples <- which(samplemissing <= 5) 
imputed.CRC.prot <- t(apply(CRC.prot[selectedSamples,], 1, function(x) random.imp(x)))
imputed.CRC.anno <- CRC.anno[selectedSamples,]
```

# 2. Heatmaps
```{r}
################################################################
################################################################
####################   H E A T M A P S  ########################
#           Iconic graphics for high-throughput data
################## CAUTION: HIDDEN PITFALLS ####################
################################################################
################################################################

################################################################
# Heatmap of individual values of proteins
################################################################
heatmap(t(imputed.CRC.prot))
?heatmap


```

## Note 2.1: choice of the distance metric matters

```{r, message=FALSE}
library(bioDist)

# toy example: 3 genes, 5 samples
gene1 <- c(1, 6, 2, 4, 7)
gene2 <- gene1 + 4 
gene3 <- gene2/3+ c(0, 2, 0, 4, 0)
e <- rbind(gene1, gene2, gene3)
dimnames(e) <- list(paste("gene", 1:3, sep=""), paste("sample", 1:5, sep=""))
e

# plot
plot(e[1,], col="red", type="l", ylim=c(0,15), xlab="samples", ylab="abundance",
     main="Original abundance values", lwd=3)
lines(e[2,], col="blue", lwd=3)
lines(e[3,], col="green", lwd=3)
legend("topright", col=c("red", "blue", "green"), paste("gene", 1:3, sep=""), lty=1, lwd=3, cex=1.5)


# default: distance fnction eucledian
heatmap(e, distfun=euc, margins=c(10,10), main="Eucledian distance")
# correlation
heatmap(e, distfun=cor.dist, margins=c(10,10), main="Correlation distance")
```


## Note 2.2: Clustering genes

genes are multivariate objects (=observations) in the space of samples (=variables)
if we are not interested in similarity of abundance 
but in similarity of the direction of change, 
then it will help to standardize the observations across variables 

now each gene has mean 0 and sample variance 1 across arrays 

```{r, message=FALSE}
library(genefilter)
e_centerScale <- ( e - rowMeans(e) ) / rowSds(e)

# plot
plot(e_centerScale[1,], col="red", type="l", ylim=c(-2,2), xlab="samples", ylab="abundance",
     main="Original abundance values")
lines(e_centerScale[2,], col="blue")
lines(e_centerScale[3,], col="green")
legend("topright", col=c("red", "blue", "green"), paste("gene", 1:3, sep=""), lty=1)

# Eucledian distance is sensitive to centering and scaling
heatmap(e, distfun=euc, margins=c(10,10), main="Eucledian, raw")
heatmap(e_centerScale, distfun=euc, margins=c(10,10), main="Eucledian, centered and scaled")

# Correlation distance is not sensitive to centering and scaling
heatmap(e, distfun=cor.dist, margins=c(10,10), main="Correlation, raw")
heatmap(e_centerScale, distfun=cor.dist, margins=c(10,10), main="Correlation, centered and scaled")
# (note that centering and scaling of genes across samples + correlation distance
# affect the order of the samples)
# my choice is to use correlation on the original values, 
# or to compute dendrograms separately, save the order of genes or samples, and use "image" to visualize 
```

## Note 2.3: The heatmap has an option "scale" by default
However it only affects the color, and not the calculation of dendrograms.
It's usually better to make custom color definitions

```{r, message=FALSE}
library(RColorBrewer)

# scaling of color matters for unscaled data
heatmap(e, distfun=euc, margins=c(10,10), scale="none", main="Eucledian, raw")
heatmap(e, distfun=euc, margins=c(10,10), main="Eucledian, scaled")
# Conclusion: genes 1 and 2 are far from each other on the dendrogram
# (because of Eucledian distance) but have same color patterns of
# intensities (because of scaled colors)


# scaling of color irrelevant if data themselves are scaled already
heatmap(e_centerScale, distfun=euc, margins=c(10,10))
heatmap(e_centerScale, distfun=euc, margins=c(10,10), scale="none")


# Example with real dataset
# Default: raw data, scaled by row (i.e. gene)
heatmap(t(imputed.CRC.prot))

# change color code: non-standardized expressions
# equal-spaced breaks for values of abundance
summary(as.vector(t(imputed.CRC.prot))) # min ~ 6, max ~ 25
exprs_brk <- seq(from=6, to=25, length=12)
heatmap(t(imputed.CRC.prot), col=brewer.pal(length(exprs_brk)-1, "RdBu"), breaks=exprs_brk)

# change color code: gene-standardized expressions
# quantile-spaced breaks in positive and negative directions
imputed_CRC_prot_centerScale <- 
 ( t(imputed.CRC.prot) - rowMeans(t(imputed.CRC.prot)) ) / rowSds(t(imputed.CRC.prot))
exprs_brk <- c( quantile(imputed_CRC_prot_centerScale[imputed_CRC_prot_centerScale<0], seq(0, 1, length=5)),
                0, quantile(imputed_CRC_prot_centerScale[imputed_CRC_prot_centerScale>0], seq(0, 1, length=5)))

heatmap(imputed_CRC_prot_centerScale, col=brewer.pal(length(exprs_brk)-1, "RdBu"), breaks=exprs_brk)
```

## Note 2.4: Label sample types

```{r, message=FALSE}
myColor <- rep("blue", nrow(imputed.CRC.prot))
myColor[imputed.CRC.anno$Sub_group == "CRC"] <- "red" 
myColor[imputed.CRC.anno$Sub_group == "Healthy"] <- "yellow" 
heatmap(imputed_CRC_prot_centerScale, col=brewer.pal(length(exprs_brk)-1, "RdBu"), breaks=exprs_brk, ColSideColors=myColor)
```


## Note 2.5: The choice of the linkage method in clustering matters
```{r, message=FALSE}
# clusters in the sample space: note the use of (t)
# method "complete" by default
hc <- hclust( dist( t(imputed_CRC_prot_centerScale)) )
plot(hc)

# sinle linkage
hc <- hclust( dist( t(imputed_CRC_prot_centerScale)), metho="single" )
plot(hc)

# cut the tree at a specific group
hc <- hclust( dist( t(imputed_CRC_prot_centerScale)) )
cc <- cutree(hc, k=2)
cc
```
