---
title: "May institute Data visualization module: Hands-on 2"
author: "Ting Huang"
date: "May 8, 2017"
output: 
  html_document:
    self_contained: true
    toc: true
    toc_float: true
    fig_caption: no	
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Summary
* Creating a new RStudio project
* Reading in data in R
* Improve figure clarity
    + Deal with overplotting
    + Order bar chart
    + Zooming
    + Change labels, themes, and scales
    
http://sebastianraschka.com/Articles/heatmaps_in_r.html

***

# 1. Create a new Rstudio project

From the menu, select **File > New Project...**, 
then select **Existing Directory** and choose the directory where you downloaded this script and the example datasets for this tutorial. 
All the output files we'll be creating in this tutorial will be saved in the 'working directory' that now has been set by Rstudio.

***

# 2. Reading in data
In this section we’ll continue using mpg and CRC dataset. We also use another diamond dataset: ~54,000 round diamonds from http://www.diamondse.info/

```{r, echo=T, warning=FALSE}
library(ggplot2)
library(RColorBrewer)
library(Biobase)

CRC <- read.csv("./data/CRC_train.csv")
CRC.prot <- CRC[,1:72]
CRC.anno <- CRC[,73:79]

# Deal with missing value
# First option: remove the samples with missing values
dim(na.omit(CRC.prot))

# Second option: impute the missing values
median.imp <- function (a){
  missing <- is.na(a)
  n.missing <- sum(missing)
  a.obs <- a[!missing]
  imputed <- a
  imputed[missing] <- sample(a.obs, n.missing, replace=TRUE)
  return (imputed)
}
pMiss <- function(x){sum(is.na(x))/length(x)*100}
samplemissing <- apply(CRC.prot,1,pMiss)
# Only keep the samples with less than 5% missing values
selectedSamples <- which(samplemissing <= 5) 
imputed.CRC.prot <- t(apply(CRC.prot[selectedSamples,], 1, function(x) median.imp(x)))
imputed.CRC.anno <- CRC.anno[selectedSamples,]

#############################################
## 2. PCA
#############################################
### 2.1 Principle components

# prcomp - based on singular value decomposition - preferred
?prcomp

# Goal: visualize samples in a space of proteins
# Therefore samples are observations; proteins are dimensions
pc <- prcomp(imputed.CRC.prot, center=TRUE, scale.=TRUE)
summary(pc)
# parts of the output
names(pc)

# proportion of explained variance
# look up the description of 'sdev' in the help of prcomp
barplot( pc$sdev^2/sum(pc$sdev^2) , xlab="principle component", 
         ylab="% of variance")
barplot( cumsum( pc$sdev^2/sum(pc$sdev^2) ) , xlab="principle component", 
         ylab="cumulative % of variance" )
# Conclusion: the two first principle components are insufficient         

# define a color for each sample 
myColor <- rep("blue", nrow(imputed.CRC.prot))
myColor[imputed.CRC.anno$Group == "CRC"] <- "red" 

# score plot
plot(pc$x[,1:2], col=myColor, pch=16, cex=2)
legend("topleft", pch=16, col=c("red", "blue"), c("CRC", "Control"))
# Conclusion: not enough separation in the first two dimensions       


### 2.2 Loadings versus differential expression
library(limma)

# Specify the design matrix X - consists of fixed factors only
design <- model.matrix(~0+imputed.CRC.anno$Group)
colnames(design) <- c("CRC", "Healthy")

# Step1: linear model. lmFit is a wrapper around lm in R
fit1 <- lmFit(t(imputed.CRC.prot), design)

# Step 2: fit contrasts: find proteins that respond to CRC cancer
contrast.matrix <- makeContrasts(CRC-Healthy, levels = design)
fit2 <- contrasts.fit(fit1, contrast.matrix)

# Step3: add empirical Bayes moderation
fit3 <- eBayes(fit2)

# adjusted p-values
p.bh <- p.adjust(fit3$p.value[,1], method = "BH")
head(p.bh)

# loadings vs adjusted p-values
smoothScatter(pc$rotation[,1], p.bh, xlab="loading", 
              ylab="adjusted p-value")
# Conclusion: large loadings may or may not correspond to small p-values

# 2.3 Principle components on a subset of significant genes
# only select differentially expressed genes
CRC_sign <- t(imputed.CRC.prot)[p.bh < 0.1,]
dim(CRC_sign)

# principle component
pc_sign <- prcomp(x=t(CRC_sign), center=TRUE, scale.=TRUE)

# score plot
plot(pc_sign$x[,1:2], col=myColor, pch=16, cex=2)
legend("topleft", pch=16, col=c("red", "blue"), c("BCR/ABL", "NEG"))
# Conclusion: better separation when only use differentially expressed genes
# (However cannot do this in practice!)

# proportion of explained variance
barplot( cumsum( pc_sign$sdev^2/sum(pc_sign$sdev^2) ) , xlab="principle component", 
         ylab="cumulative % of variance" )
# Conclusion: two PC are still not enough in this case         

# loadings vs adjusted p-values
# Conclusion: better correspondence, but the loading cutoff 
# and control of FDR are unclear
smoothScatter(pc_sign$rotation[,1], p.bh[p.bh < 0.1], xlab="loading", ylab="adjusted p-value")


# OVERALL CONCLUSION:
# Do not use loadings as evidence of differential expression

#############################################
## 3. Scatterplot matrices
#############################################
# Reference: https://gastonsanchez.wordpress.com/2012/08/27/scatterplot-matrices-with-ggplot/
# install.packages("GGally")
pairs(CRC[,c("CP", "PON1", "SERPINA3", "LRG1", "TIMP1")])

library(GGally)
ggpairs(CRC[,c("CP", "PON1", "SERPINA3", "LRG1", "TIMP1", "Sub_group")], aes(colour = Sub_group, alpha = 0.4), upper = list(continuous = wrap("cor", size = 3)))
pairs(CRC[,c("CP", "PON1", "SERPINA3", "LRG1", "TIMP1")])
#############################################
## 4. Venn diagram
#############################################
# Reference: https://cran.r-project.org/web/packages/UpSetR/vignettes/basic.usage.html
# install.packages("VennDiagram")
library(VennDiagram)
A <- sample(1:1000, 400, replace = FALSE); 
B <- sample(1:1000, 600, replace = FALSE);
C <- sample(1:1000, 350, replace = FALSE);
D <- sample(1:1000, 550, replace = FALSE);
E <- sample(1:1000, 375, replace = FALSE);
venn.diagram(x = list(A = A, D = D, B = B, C = C), filename = "Venn.tiff",
             col = "transparent", fill = c("cornflowerblue","green","yellow","darkorchid1"),
             alpha = 0.50, label.col = c("orange", "white", "darkorchid4", "white", "white", 
                                         "white",    "white", "white", "darkblue", "white", "white", "white", "white", 
                                         "darkgreen", "white"), cex = 1.5, fontfamily = "serif", fontface = "bold",
             cat.col = c("darkblue", "darkgreen", "orange", "darkorchid4"), cat.cex = 1.5,
             cat.pos = 0, cat.dist = 0.07, cat.fontfamily = "serif", rotation.degree = 270,
             margin = 0.2)

# UpSetR Package
# install.packages("UpSetR") 
library(UpSetR)
listInput <- list(A = A, D = D, B = B, C = C)
upset(fromList(listInput), order.by = "freq")
```
