---
title: "May institute Data visualization module: Hands-on 2"
author: "Ting Huang"
date: "May 8, 2017"
output: 
  html_document:
    self_contained: true
    toc: true
    toc_float: true
    fig_caption: no	
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Summary

* Reading in data in R
* PCA
    + Principle components
    + Loadings versus differential expression
    + Principle components on a subset of significant genes
* Scatterplot matrices
* Venn diagram

---

# 1. Reading in data
In this section we’ll continue using mpg and CRC dataset. We also use another diamond dataset: ~54,000 round diamonds from http://www.diamondse.info/

```{r, echo=T, warning=FALSE}
library(ggplot2)
library(RColorBrewer)
library(Biobase)

CRC <- read.csv("./data/CRC_train.csv")
CRC.prot <- CRC[,1:72]
CRC.anno <- CRC[,73:79]

# Deal with missing value
# First option: remove the samples with missing values
dim(na.omit(CRC.prot))

# Second option: impute the missing values
median.imp <- function (a){
  missing <- is.na(a)
  n.missing <- sum(missing)
  a.obs <- a[!missing]
  imputed <- a
  imputed[missing] <- sample(a.obs, n.missing, replace=TRUE)
  return (imputed)
}
pMiss <- function(x){sum(is.na(x))/length(x)*100}
samplemissing <- apply(CRC.prot,1,pMiss)
# Only keep the samples with less than 5% missing values
selectedSamples <- which(samplemissing <= 5) 
imputed.CRC.prot <- t(apply(CRC.prot[selectedSamples,], 1, function(x) median.imp(x)))
imputed.CRC.anno <- CRC.anno[selectedSamples,]
```

---

#############################################
# 2. PCA
#############################################
## 2.1 Principle components
```{r}
# prcomp - based on singular value decomposition - preferred
?prcomp

# Goal: visualize samples in a space of proteins
# Therefore samples are observations; proteins are dimensions
pc <- prcomp(imputed.CRC.prot, center=TRUE, scale.=TRUE)
summary(pc)
# parts of the output
names(pc)

# proportion of explained variance
# look up the description of 'sdev' in the help of prcomp
pc.var <- data.frame(index = c(1:length(pc$sdev)), var.prop = pc$sdev^2/sum(pc$sdev^2), var.prop.cum = cumsum( pc$sdev^2/sum(pc$sdev^2)))
ggplot(pc.var) +
  geom_bar(aes(index, var.prop), stat = "identity") + 
  labs(title = "PCA",
        x = "principle component",
        y = "% of variance")

ggplot(pc.var) +
  geom_bar(aes(index, var.prop.cum), stat = "identity") + 
  labs(title = "PCA",
        x = "principle component",
        y = "cumulative % of variance")

# Conclusion: the two first principle components are insufficient         

# define a color for each sample 
myColor <- rep("blue", nrow(imputed.CRC.prot))
myColor[imputed.CRC.anno$Group == "CRC"] <- "red" 

# scatter plot
PCs <- as.data.frame(pc$x)
PCs$Group <- imputed.CRC.anno$Group
g <- ggplot(PCs, aes(x = PC1, y = PC2, color = Group))
g + geom_point()

# Conclusion: not enough separation in the first two dimensions       
```

## 2.2 Loadings versus differential expression
```{r}
library(limma)
library(ggrepel)
# Specify the design matrix X - consists of fixed factors only
design <- model.matrix(~0+imputed.CRC.anno$Group)
colnames(design) <- c("CRC", "Healthy")

# Step1: linear model. lmFit is a wrapper around lm in R
fit1 <- lmFit(t(imputed.CRC.prot), design)

# Step 2: fit contrasts: find proteins that respond to CRC cancer
contrast.matrix <- makeContrasts(CRC-Healthy, levels = design)
fit2 <- contrasts.fit(fit1, contrast.matrix)

# Step3: add empirical Bayes moderation
fit3 <- eBayes(fit2)

# adjusted p-values
# Benjamini-Hochberg (FDR) test for multiple correction
p.bh <- topTable(fit3, adjust = "BH", sort = "none", n = Inf)
p.bh$Protein <- rownames(p.bh)

p.bh$Significant <- ifelse(p.bh$adj.P.Val < 0.05, "FDR < 0.05", "Not Sig")
ggplot(p.bh, aes(x = logFC, y = -log10(P.Value))) +
  geom_point(aes(color = Significant)) +
  scale_color_manual(values = c("red", "grey")) +
  theme_bw() +
  geom_text_repel(
    data = subset(p.bh, adj.P.Val < 0.05),
    aes(label = Protein)
  )

# loadings vs adjusted p-values
smoothScatter(pc$rotation[,1], p.bh$adj.P.Val, xlab="loading", 
              ylab="adjusted p-value")
```
Conclusion: large loadings may or may not correspond to small p-values

## 2.3 Principle components on a subset of significant genes
We only select differentially expressed genes

```{r}
CRC_sign <- t(imputed.CRC.prot)[p.bh$adj.P.Val < 0.1,]
dim(CRC_sign)

# principle component
pc_sign <- prcomp(x=t(CRC_sign), center=TRUE, scale.=TRUE)

# scatter plot
PCs_sign <- as.data.frame(pc_sign$x)
PCs_sign$Group <- imputed.CRC.anno$Group
g <- ggplot(PCs_sign, aes(x = PC1, y = PC2, color = Group))
g + geom_point()

# Conclusion: better separation when only use differentially expressed genes
# (However cannot do this in practice!)

# proportion of explained variance
barplot( cumsum( pc_sign$sdev^2/sum(pc_sign$sdev^2) ) , xlab="principle component", 
         ylab="cumulative % of variance" )
# Conclusion: two PC are still not enough in this case         

# loadings vs adjusted p-values
# Conclusion: better correspondence, but the loading cutoff 
# and control of FDR are unclear
smoothScatter(pc_sign$rotation[,1], p.bh$adj.P.Val[p.bh$adj.P.Val < 0.1], xlab="loading", ylab="adjusted p-value")
```

OVERALL CONCLUSION:
Do not use loadings as evidence of differential expression

---

#############################################
# 3. Scatterplot matrices
#############################################
Reference: https://gastonsanchez.wordpress.com/2012/08/27/scatterplot-matrices-with-ggplot/
```{r}
pairs(CRC[,c("CP", "PON1", "SERPINA3", "LRG1", "TIMP1")])
# install.packages("GGally")
library(GGally)

ggpairs(CRC[,c("CP", "PON1", "SERPINA3", "LRG1", "TIMP1", "Sub_group")], aes(colour = Sub_group, alpha = 0.4), upper = list(continuous = wrap("cor", size = 3)))
pairs(CRC[,c("CP", "PON1", "SERPINA3", "LRG1", "TIMP1")])
```

#############################################
# 4. Venn diagram
#############################################
Reference: https://cran.r-project.org/web/packages/UpSetR/vignettes/basic.usage.html

```{r}
# install.packages("VennDiagram")
library(VennDiagram)
A <- sample(1:1000, 400, replace = FALSE); 
B <- sample(1:1000, 600, replace = FALSE);
C <- sample(1:1000, 350, replace = FALSE);
D <- sample(1:1000, 550, replace = FALSE);
E <- sample(1:1000, 375, replace = FALSE);
venn.diagram(x = list(A = A, D = D, B = B, C = C), filename = "Venn.tiff",
             col = "transparent", fill = c("cornflowerblue","green","yellow","darkorchid1"),
             alpha = 0.50, label.col = c("orange", "white", "darkorchid4", "white", "white", 
                                         "white",    "white", "white", "darkblue", "white", "white", "white", "white", 
                                         "darkgreen", "white"), cex = 1.5, fontfamily = "serif", fontface = "bold",
             cat.col = c("darkblue", "darkgreen", "orange", "darkorchid4"), cat.cex = 1.5,
             cat.pos = 0, cat.dist = 0.07, cat.fontfamily = "serif", rotation.degree = 270,
             margin = 0.2)

# UpSetR Package
# install.packages("UpSetR") 
library(UpSetR)
listInput <- list(A = A, D = D, B = B, C = C)
upset(fromList(listInput), order.by = "freq")
```

---