---
title: "Programming with Functions"
author: "Jan Vitek"
date: "Wednesday May 3, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{echo=F}
Learning goals
- Basics: what is a function and why would you write one?
- How to write a function
- Parts of a function
    + Formals
    + Body
    + Environment
- Example: writing a function
    + Rewrite a previous example as a function
- Function arguments
    + Default and missing arguments
    + Lazy evaluation
- Returning values: implicit, `return()`, `invisible()`
- John Chambers: "To understand computations in R, two slogans are helpful:"
    + *Everything that exists is an object*
    + *Everything that happens is a function call*
- Example: `+` and `-`, etc., are also functions, can be redefined
- Infix functions, creating a user-defined `%plus%`, etc.
- How to write good functions
```

Code often grows organically without structure or discipline. Code often is the result of a sequence of cut-and-paste-with-replacement operations.

```{r}
x = matrix(1:9, nrow=3,ncol=3)
r = 0
r[1] = 10 + x[1,1] * 2 + x[1,2] / 2 + x[1,3] / 4
r[1] = r[1]*r[1]
if (is.na(x[2,3]) == FALSE) {
  r[2] =  x[2,1] * 2 + x[2,2] / 2 + x[2,3] / 4 + 100
} else {
  r[2] = NA
}
r[3] = x[3,1] * 2 + x[3,2] / 2 + x[3,3] / 4 + 1000
r[3] = r[3]*r[3]
r[2] = r[2]* r[2]
r[3] = as.integer(r[3])
r[2] = as.integer(r[2])
r[1] = as.integer(r[1])
```
  

*Exercise* assume that the above code was something that needed to be reused. Design a functional abstraction to encapsulate it cleanly.


Laws:
  
  - code is read more often than written: optimize for the readers

  - avoid repetition: if two code snippets do the same thing, turn them into a function
  
  - untested code is wrong code
  
   - do not repeat yourself or DRY, principle.   -- from "pragmatic programmers", Dave Thomas and Andy Hunt:
       "every piece of knowledge must have a single, unambiguous, authoritative representation within a system"
  

```{r}
set.seed(1014)
df <- data.frame(replicate(6, sample(c(1:10, -99), 6, rep = TRUE)))
names(df) <- letters[1:6]
df$a[df$a == -99] <- NA
df$b[df$b == -99] <- NA
df$c[df$c == -98] <- NA
df$d[df$d == -99] <- NA
df$e[df$e == -99] <- NA
df$f[df$g == -99] <- NA
```

*Exercise* rewrite the above with a function to replace the missings.


## Functions

The building block for creating abstractions in R is the *function declaration*.  These abstractions are used in *function calls*.

A function object is a value in the 

```{r, eval=FALSE}
f <- function( y ) {  y + 1 }
f( 1 )
```

Function definitions in R consist of an argument list and a body. The argument list is a sequence of variable names (with optional default values) and possibly one occurrence of the tripe dot symbol (`...`).  The body of the function can be a single expression or a block (`{...}`).

```{r, eval=FALSE}
# the shortest function in the book
function() 1    
# multiple expressions in the body
function( x ) { y <- x * x;  y + 1 }
# default values, optional arguments
function( x = 0, ..., y = 2)   if(length(z <- list(...) == 0)) x * y  else x * y * z[[1]] 
```

A function call consists of an expression that evaluates to a function followed by a (possibly empty) list of (possible named) arguments.

Functions can be stored in lists, e.g. `list(f)`.

```{r, eval=FALSE}
f()
# the shortest call of them all
f(1,2)
# positional calls pass arguments in order
f(x = 1, y = 2)x`
f(y = 2, x = 1)
# named calls pass argument by name without a prescribed order
(x[42])(x=1)
# an expression that yields a function can be called
```

**Exercise** write a definition of a function `f` that will return `0`, `-1`, `1`, `1` for the four calls to `f` above and a variable `x` such that the last call returns `-1`.



Some functions have inline syntax, `+` for example does not require parentheses and `if else` is  function which takes two arguments.  The assignment operators are functions `<-` and `[]<-, so are parenthese `(`.  All functions can be redefined.


To define an inline operator, one can either redefine an existing operator like `+` or use the syntax `%+%` to define a new operator.

```{r}
"%+%" <- function(a,b) paste(a,b)

"hi" %+% "ho"
```

This allows users to extend the set of operators but with the risk of collisions.



```{r, eval=F}
f <- function(y = 0, x = 0) x - y
f()
f(1,2)
f(x = 1, y = 2)
f(y = 2, x = 1)
```

The optional arguments have somewhat interesting semantics. In a function, any argument occurring after the three dots must be passed by name.  In the function body they can be passed as argument to a further call.

Function evaluation is complete when the last expression in the body is evaluated. The `return()` call can break evaluation in the middle.  The `stop()` call can raise an exception that can only be caught in a `tryCatch()` statement. 

```{r, eval=F}
f <- function(x,acc) {
  if (x == 0) tryCatch( stop(simpleException("done")), 
                        simpleException = function(err) acc)
  else f(x-1, acc+1) 
}  
```


Each call to a function adds the argument to function on a data structure referred to as the *stack*. When a function returns the arguments are removed from the stack.

### Referential transparency

To facilitate equational reasoning, R attempts to provide *referential transparency* for function calls. Referential transparency means that arguments are not changed by the function being called. So the following function `f(x)` does not modify the vector passed in.

```{r}
x
f <- function(x) { x[1] <- 1 }
f(x)
x
```

But then what does it change in the body?

What are the statements that cause this happen?


## Delayed Evaluation

**Exercise** Consider the following code, what will it print?

```{r}
print(1)   # outputs "1" on the terminal
f<-function(a,b) { invisible(0) }  # a helper, nevermind
f( print(1) , print(2) )   # what would this do in Java or in C
temp <-print(1);temp2 <- print(2); f(temp, temp2)  # same thing?
```

**Exercise** What will this print?

```{r}
f1 <- function( a, b, c ) { # Consider the following, not very useful function
 x <- f(b,b)
 x <- c
 x <- c
 a
}

f1( print(1), print(2), print(3) ) # What do you recon this will print?
```

**Exercise** What will this print?

```{r}
f2 <- function( a, b, c ) { # here is another function
  x <- b
  x <- c
  f3(a)
}

f3 <- function( a) { } # with a silly helper function

f2( print(1), print(2), print(3) ) # what will we print?
```

In R, arguments to a functions are not evaluated, they are packaged into *promises*. A promise is only evaluated if it is accessed (and only the first time it is accessed).  

Every expression is passed as a promise, except constants.

The following is an obscure corner of R.   Enjoy.

```{r}
FU <- function(b="") print(paste("FU",b))                # A function
BAR <- function( FU = print("BAR"))  FU("bar")   # Another function with a default argument
BAR()   # call bar will print ?
BAR( function(n) print("YO") )  # and now?
```


## Building a counter abstraction

With referential transparency it is difficult to mutate non-local state.  What if we wanted to create a large number of counters and did not want anyone but us to be able to update them.



```{r}
remove(xPriv)
counter <- function() { xPriv <- xPriv + 1; xPriv}
#counter()             # oops, we need to define x
xPriv<-0
counter()
counter()             # Hmm... not quite

# Super assignment to the rescue
counter <- function() { xPriv <<- xPriv + 1; xPriv}  # <<- assigns in the parent env
counter()
counter()             # now we have a counter
```

This only lets us have a single counter. What if we want a counter factory?


```{r}
mkC <- function () { 
    xPriv <- 0
    function() { xPriv <<- xPriv + 1; xPriv }
  }

count <- mkC()
count()
count()
```

The above uses the environment of the `mkC()` function to store the counter. Each time we call `mkC`, a new environment is created.


## Build abstraction with domain specific languages

R allows to leverage delayed evaluation in order to implement domain specific languages.

Here is the example of timer function that repeatedly run the expression passed in as argument and reports the timing of that expression.

```{r}
tm <- function( exp ) {
 x<- deparse(substitute( exp ))   ## decompile the expression
 t <- 1                           ## repetion counter (we run exp 10 * t)
 time = 0;                        ## reset timer
 while ( time < 1 ) {             ## continue looking until the code runs long enough
  y<- paste(" system.time(for(i in 1:",t,"){",x,"})")
  r <- eval(parse(text=y))        ## execute
  time <- r[3]                    ## grab the elasped time
  t <- t * 10                     ## try with a bigger t
 }
 (time / t) * 1000                ## return time in milliseconds (default is in s)
}
tm(1 + 1)
```


# Testing 

```{r}
#install.packages ("testthat")
library(stringr)
library(testthat)

test_that("str_length is number of characters", {
  expect_equal(str_length("a"), 1)
  expect_equal(str_length("ab"), 2)
  expect_equal(str_length("abc"), 3)
})

test_that("str_length of factor is length of level", {
  expect_equal(str_length(factor("a")), 1)
  expect_equal(str_length(factor("ab")), 2)
  expect_equal(str_length(factor("abc")), 3)
})

test_that("str_length of missing is missing", {
  expect_equal(str_length(NA), NA_integer_)
  expect_equal(str_length(c(NA, 1)), c(NA, 1))
  expect_equal(str_length("NA"), 2)
})
```

